Baseball

### 2024-02-15 example_projects/baseball/docs/report.qmd
Update baseball example project

---
title: "Modelling at-bats in baseball using the generalised Pareto distribution"
subtitle: "A report created using bibat version 0.1.6"
author: "Teddy Groves"
bibliography: bibliography.bib
reference-location: margin
toc: true
format:
  html:
    embed-resources: true
filters:
  - include-code-files
---

@@

Even though I don't work with sports data very much any more, the problem still
haunted me, so when I read [this great case
study](https://mc-stan.org/users/documentation/case-studies/gpareto_functions.html#conclusion-on-the-data-analysis)
about geomagnetic storms it gave me an idea for yet another potential solution.

The idea was this: just as data about intense storms that cause electricity
problems tell us about a tail of the bigger solar magnetism distribution, maybe
data about professional sportspeople is best thought about as coming from a tail
of the general sports ability distribution. If so, maybe something like the
[generalised pareto
distribution](https://en.wikipedia.org/wiki/Generalized_Pareto_distribution)
might be better than the bog standard normal distribution for describing the
pros' abilities.

@@

With a few datasets and at least two different statistical models to consider,
the full analysis looked like it would be a little too big to fit in a trivial
file structure, so it seemed like a good opportunity to showcase my
batteries-included Bayesian analysis template package
[bibat](https://github.com/teddygroves/bibat).

@@

Check out the analysis's [github
repository](https://github.com/teddygroves/baseball) for all the details.

@@

First I installed bibat in my global Python 3.11 environment with this command:

```zsh
> pip install bibat
```

@@

```{.zsh}
> bibat
Welcome to the Batteries-Included Bayesian Analysis Template!
What is your project called? [Project Name]: Baseball
What should the project repository be called [baseball]: baseball
What is your name? [Author name]: Teddy Groves
What is your email? [author@email.com]:
Who should be the code of conduct contact? [author@email.com]:
Please briefly describe your project [A short description of the project.]: Comparison of distributions for modelling baseball hitting
Choose an open source license from these options: (MIT, BSD-3-Clause, No license file) [MIT]:
How would you like to document your project? (Quarto, Sphinx, No docs) [Quarto]:
Would you like to create a .github directory? [y]: n
>
```

@@

```{.zsh}
baseball
├── CODE_OF_CONDUCT.md
├── LICENSE
├── Makefile
├── README.md
├── bibat_version.txt
├── data
│   └── raw
│       ├── raw_measurements.csv
│       └── readme.md
├── docs
│   ├── bibliography.bib
│   ├── img
│   │   ├── example.png
│   │   └── readme.md
│   ├── report.html
│   ├── report.pdf
│   └── report.qmd
├── baseball
│   ├── __init__.py
│   ├── data_preparation.py
│   ├── fitting_mode.py
│   ├── inference_configuration.py
│   ├── investigate.ipynb
│   ├── notebooks
│   ├── readme.md
│   ├── sample.py
│   ├── stan
│   │   ├── custom_functions.stan
│   │   ├── multilevel-linear-regression.stan
│   │   └── readme.md
│   ├── stan_input_functions.py
│   └── util.py
├── inferences
│   ├── fake_interaction
│   │   └── config.toml
│   ├── interaction
│   │   └── config.toml
│   └── no_interaction
│       └── config.toml
├── plots
│   ├── posterior_ll_comparison.png
│   └── posterior_predictive_comparison.png
├── pyproject.toml
└── tests
    ├── test_integration
    │   └── test_data_preparation.py
    └── test_unit
        ├── test_inference_configuration.py
        └── test_util.py
```

This folder implements bibat's example analysis – a comparison of linear
regression with two different design matrices. To check that everything was
working correctly I ran the analysis like this:

@@

```{.python include="../baseball/fetch_data.py"}
```

@@

```zsh
> source .venv/bin/activate
(baseball) > python baseball/fetch_data.py
Fetching 2006 data from https://raw.githubusercontent.com/stan-dev/example-models/master/knitr/pool-binary-trials/baseball-hits-2006.csv
Writing 2006 data to data/raw/2006.csv
Fetching bdb-main data from https://raw.githubusercontent.com/cbwinslow/baseballdatabank/master/core/Batting.csv
Writing bdb-main data to data/raw/bdb-main.csv
Fetching bdb-post data from https://raw.githubusercontent.com/cbwinslow/baseballdatabank/master/core/BattingPost.csv
Writing bdb-post data to data/raw/bdb-post.csv
Fetching bdb-apps data from https://raw.githubusercontent.com/cbwinslow/baseballdatabank/master/core/Appearances.csv
Writing bdb-apps data to data/raw/bdb-apps.csv
```

@@

```zsh
(baseball) > rm data/raw/raw_measurements.csv
```

# Preparing the data

The first step in preparing data is to decide what prepared data looks like for
the purposes of our analysis. Bibat provides dataclasses called `PreparedData`
and `MeasurementsDF` to help get started with this, which I found in the file
`baseball/data_preparation.py`.

As it happens, prepared data looks very similar in this analysis and the
example. All I had to do was change the `MeasurementsDF` definition a
little^[note that this class uses [pandera](https://pandera.readthedocs.io), a
handy library for defining what a pandas dataframe should look like]:

```python
class MeasurementsDF(pa.SchemaModel):
    """A PreparedData should have a measurements dataframe like this.

    Other columns are also allowed!
    """

    player_season: Series[str]
    season: Series[str] = pa.Field(coerce=True)
    n_attempt: Series[int] = pa.Field(ge = 1)
    n_success: Series[int] = pa.Field(ge = 0)
```

The next step is to write functions that return `PreparedData` objects. In this
case I wrote a couple of data preparation functions: `prepare_data_2006` and
`prepare_data_bdb`:

```{.python include="../baseball/data_preparation.py" start-line=74 end-line=159}
```

To take into account the inconsistency between the two raw data sources, I
first had to change the variable `RAW_DATA_FILES`:

```{.python include="../baseball/data_preparation.py" start-line=26 end-line=33}
```

@@

```{.python include="../baseball/data_preparation.py" start-line=25 end-line=32}
```

To finish off I deleted the unused global variables `NEW_COLNAMES`,
`DROPNA_COLS` and `DIMS`, then checked if the function `load_prepared_data`
needed any changes: I was pretty sure it didn't.

@@

Now the folder `data/prepared/bdb` contained a file
`data/prepared/bdb/measurements.csv` that looked like this:

```{csv}
,player_season,season,n_attempt,n_success
0,abreujo02,2018,553,180
1,acunaro01,2018,487,178
3,adamewi01,2018,322,112
6,adamsla01,2018,29,10
7,adamsma01,2018,337,104
8,adamsma01,2018,277,92
9,adamsma01,2018,60,12
```

@@

```{.stan include="../baseball/stan/gpareto.stan"}
```

@@

```{.stan include="../baseball/stan/normal.stan"}
```

@@

Next I needed to tell the analysis how to turn some prepared data into a
dictionary that can be used as input for Stan. Bibat assumes that this task is
handled by functions that live in the file `baseball/stan_input_functions.py`,
each of which takes in a `PreparedData` and returns a Python dictionary. You
can write as many Stan input functions as you like and choose which one to run
for any given inference.

I started by defining some Stan input functions that pass arbitrary prepared
data on to each of the models:^[Note that this code uses the scipy function
`logit`, which it imported like this: `from scipy.special import logit`]

```{.python include="../baseball/stan_input_functions.py" start-line=11 end-line=34}
```

But why stop there? It can also be useful to generate Stan input data with a
model, by running it in simulation mode with hardcoded parameter values. Here
are some functions that do this for both of our models:
^[These functions require some more imports: `numpy` and `scipy.special.expit`]
```{.python include="../baseball/stan_input_functions.py" start-line=36}
```

@@

Now all the building blocks for making statistical inferences – raw data, data
preparation rules, statistical models and recipes for turning prepared data
into model inputs – were in place. The last step before actually running Stan
was to write down how put the blocks together. Bibat has another concept for
this, called 'inferences'.

@@

- The Stan file, prepared data folder and Stan input function are referred to by
  strings. The analysis should raise an error if you enter a non-existing value.
- Both inferences are set to run in "prior" and "posterior" modes – the other
  pre-defined mode is "kfold", but you can also write your own!
- You can enter arbitrary arguments to cmdstanpy's `CmdStanModel.sample` method
  in the `[sample_kwargs]` table.
- You can enter mode-specific overrides in `[mode_options.<MODE>]`. This can be
  handy if you want to run more or fewer iterations for a certain mode.

@@

```zsh
inferences
├── gpareto2006
│   ├── config.toml
│   └── idata.json
└── normal2006
    ├── config.toml
    └── idata.json
```

@@

Now that the inferences are ready it's time to check them out. Bibat provides a
Jupyter notebook at `baseball/investigate.ipynb` for exactly this purpose. The
notebook's main job is to create plots and save them in the `plots` directory
when it is executed with the command `jupyter execute investigate.ipynb`, which
is the final step in the chain of commands that is triggered by `make
analysis`.

@@

The code from the example analysis's notebook for loading `InferenceData` was
reusable with a few tweaks to avoid missing file errors, so I kept it. On the
other hand, I wanted to make some different plots from the ones in the example
analysis, including some that required loading prepared data. To check out
everything I did, see
[here](https://github.com/teddygroves/bibat/blob/main/bibat/examples/baseball/baseball/investigate.ipynb).

@@

To model the more recent data, all I had to do was create some new inference
folders with appropriate `prepared_data_dir` fields in their `config.toml`
files. For example, here is the `config.toml` file for the `gparetobdb`
inference:

@@

After running `make analysis` one more time, I went back to the notebook
`baseball/investigate.ipynb` and made plots of both models' posterior 1%-99%
success rate intervals for both datasets:

@@

The final step was to document my analysis. To do this I edited the file
`docs/report.qmd`, then ran `make docs`, which produced the
very HTML document that you are probably reading now! You can find the complete
`report.qmd` file
[here](https://github.com/teddygroves/bibat/blob/main/bibat/examples/baseball/docs/report.qmd).

>>>>

---
title: "Baseball"
resource_path: ["../"]
subtitle: "A report created using bibat version 0.3.0"
author: "Teddy Groves"
bibliography: bibliography.bib
reference-location: margin
toc: true
format:
  html:
    embed-resources: true
  pdf: default
filters:
  - include-code-files
---

@@

Even though I don't work with sports data very much any more, these issues still
lingered in my mind, so when I read [this great case
study](https://mc-stan.org/users/documentation/case-studies/gpareto_functions.html#conclusion-on-the-data-analysis)
about geomagnetic storms it gave me an idea for yet another potential solution.

The idea was this: just as data about intense storms tell us about a tail
of the bigger solar magnetism distribution, maybe data about professional
sportspeople is best thought about as coming from a tail of the general sports
ability distribution. If so, perhaps something like the [generalised pareto
distribution](https://en.wikipedia.org/wiki/Generalized_Pareto_distribution)
might be better than the bog-standard normal distribution for describing the
pros' abilities.

@@

With a few datasets and statistical models to consider, the full analysis looked
like it would be too big to fit in a single file, so this seemed like a job for a
batteries-included Bayesian analysis template like [bibat](https://github.com/teddygroves/bibat).

@@

Check out [the full analysis](https://github.com/teddygroves/bibat/example_projects/baseball) for all the details.

@@

First I installed [copier](https://copier.readthedocs.io/en/stable/) in my
global Python 3.12 environment with this command:

```zsh
> pipx install copier
```

@@

```{.zsh}
☁  teddygroves  copier copy gh:teddygroves/bibat baseballl
🎤 Name of your project
   Baseball
🎤 Name of your project, with no spaces (used for venv and package names)
   baseball
🎤 A short description of the project.
   Is the generalised Pareto distribution good for modelling latent hitting
ability?
🎤 Your name (or your organization/company/team)
   Teddy Groves
🎤 Author email (will be included in pyproject.toml)
   groves.teddy@gmail.com
🎤 Code of conduct contact
   groves.teddy@gmail.com
🎤 open_source_license
   MIT
🎤 docs_format
   Quarto
🎤 create_dotgithub_directory (bool)
   No
```

@@

```{.zsh}
☁  teddygroves  tree baseball
baseball
├── CODE_OF_CONDUCT.md
├── LICENSE
├── Makefile
├── README.md
├── data
│   └── raw
│       ├── raw_measurements.csv
│       └── readme.md
├── docs
│   ├── bibliography.bib
│   ├── img
│   │   ├── example.png
│   │   └── readme.md
│   └── report.qmd
├── inferences
│   ├── fake_interaction
│   │   └── config.toml
│   ├── interaction
│   │   └── config.toml
│   └── no_interaction
│       └── config.toml
├── notebooks
│   └── investigate.ipynb
├── plots
│   └── readme.md
├── pyproject.toml
├── src
│   ├── __init__.py
│   ├── data_preparation.py
│   ├── fitting.py
│   ├── stan
│   │   ├── custom_functions.stan
│   │   ├── multilevel-linear-regression.stan
│   │   └── readme.md
│   └── stan_input_functions.py
└── tests
    └── test_integration
        └── test_data_preparation.py
```

This folder implements bibat's example analysis, which compares linear
regression models with different design matrices. To check that everything was
working correctly I ran the analysis like this:

@@

```{.python include="../src/fetch_data.py"}
```

@@

```zsh
(baseball) ☁  baseball  python src/fetch_data.py --verbose
/Users/tedgro/repos/teddygroves/baseball/src/fetch_data.py:6: DeprecationWarning:
Pyarrow will become a required dependency of pandas in the next major release of pandas (pandas 3.0),
(to allow more performant data types, such as the Arrow string type, and better interoperability with other libraries)
but was not found to be installed on your system.
If this would cause problems for you,
please provide us feedback at https://github.com/pandas-dev/pandas/issues/54466

  import pandas as pd
INFO:root:Fetching 2006 data from https://raw.githubusercontent.com/stan-dev/example-models/master/knitr/pool-binary-trials/baseball-hits-2006.csv
INFO:root:Writing 2006 data to data/raw/2006.csv
INFO:root:Fetching bdb-main data from https://raw.githubusercontent.com/cbwinslow/baseballdatabank/master/core/Batting.csv
INFO:root:Writing bdb-main data to data/raw/bdb-main.csv
INFO:root:Fetching bdb-post data from https://raw.githubusercontent.com/cbwinslow/baseballdatabank/master/core/BattingPost.csv
INFO:root:Writing bdb-post data to data/raw/bdb-post.csv
INFO:root:Fetching bdb-apps data from https://raw.githubusercontent.com/cbwinslow/baseballdatabank/master/core/Appearances.csv
INFO:root:Writing bdb-apps data to data/raw/bdb-apps.csv
```

@@

```zsh
> rm data/raw/raw_measurements.csv
```

# Data Preparation

The example project already has some data preparation code in the file
`src/data_preparation.py`: I mostly just had to adapt what was already there.

The first data preparation step was to decide what prepared data should
look like for my analysis and represent this definition using a subclass of
bibat's `PreparedData` [Pydantic `BaseModel`](https://docs.pydantic.dev/latest/api/base_model/). Looking at the example analysis's
`ExamplePreparedData` class, I could see that it already had the components
that I needed, namely a name, a dictionary of coordinates and a table of
measurements:

```{.python include="../../../bibat/template/src/data_preparation.py" start-line=44 end-line=57}
```

There are a few interesting things about this model. First, it uses some
specialised types from bibat's `util` module, namely `CoordDict` and
`DfInPydanticModel`. The latter is quite handy as it ensures that the table of
measurements can be saved to json and then safely read back again. Second, the
measurement table has a validator that refers to a [pandera `DataFrameModel`](https://pandera.readthedocs.io/en/stable/reference/generated/pandera.api.pandas.model.DataFrameModel.html?highlight=SchemaModel) called `ExampleMeasurementsDF` that defines what a measurement table should be like. It is defined just before `ExamplePreparedData`:

```{.python include="../../../bibat/template/src/data_preparation.py" start-line=32 end-line=41}
```

I only had to change a few things to make new definitions for my analysis:

```{.python include="../src/data_preparation.py" start-line=39 end-line=61}
```

The next step is to write functions that return `BaseballPreparedData` objects.
In this case I wrote a couple of data preparation functions: `prepare_data_2006`
and `prepare_data_bdb`:

```{.python include="../src/data_preparation.py" start-line=64 end-line=157}
```

To take into account the inconsistency between my two raw data sources, I first
had to change the variable RAW_DATA_FILES:

```{.python include="../src/data_preparation.py" start-line=29 end-line=36}
```

@@

```{.python include="../src/data_preparation.py" start-line=159 end-line=177}
```

To finish off the data preparation step I deleted any unused code from the
example analysis, and updated the function `load_prepared_data`'s signature:

```{.python include="../src/data_preparation.py" start-line=180 end-line=185}
```

@@

Now the folder `data/prepared/` contained json files `2006.json` and `bdb.json`
that looked like this:

```zsh
(baseball) > jq . data/prepared/bdb.json | head -n 10
{
  "name": "bdb",
  "coords": {
    "player_season": [
      "abreujo022017",
      "abreujo022018",
      "abreujo022019",
      "abreujo022020",
      "abreujo022021",
      "acunaro012018",
```

@@

```{.stan include="../src/stan/gpareto.stan"}
```

@@

```{.stan include="../src/stan/normal.stan"}
```

@@

Next I needed to tell the analysis how to turn some prepared data into a
dictionary that can be used as input for Stan. The example analysis includes
some functions that do this in the file `src/stan_input_functions.py`.

I followed the examples to create similar functions for my two Stan
models:^[Note that this code uses the scipy function `logit`, which it imported
like this: `from scipy.special import logit`]

```{.python include="../src/stan_input_functions.py" end-line=33}
```

<!-- todo: add link for returns_stan_input below -->

These function use a handy decorator called `returns_stan_input` that uses the
power of [`stanio`](https://github.com/WardBrian/stanio) to convert a dictionary
to a json serialisable, Stan-friendly form. Thanks to this decorator I didn't
have to spend time appending `.values.tolist()` to the inputs that came from
pandas objects.

I now had functions that turned my prepared data into input for either of my
Stan models. But why stop there? It can also be useful to generate fake input
data, by running a model in simulation mode using hardcoded parameter values.
Here are some functions that do this for both of my models: ^[These functions
require some more imports: `numpy` and `scipy.special.expit`]

```{.python include="../src/stan_input_functions.py" start-line=37}
```

# Configuring fitting

To tell my analysis how to run inferences, I had to edit the file `src/
fitting.py`. The only changes I made from the example analysis were to remove
the `kfold` fitting mode, which I didn't need to use in this analysis, and to
add my Stan input functions to the `LOCAL_FUNCTIONS` constant:

```{.python include="../src/fitting.py"}
```

@@

Now all the building blocks for making statistical inferences – raw data, data
preparation rules, statistical models, recipes for turning prepared data into
model inputs and a procedure for running inferences – were in place. The last
step before actually running my analysis was to write down how put the blocks
together. Bibat has another concept for this, called 'inferences'.

@@

- The Stan file, prepared data folder and Stan input function are referred to by
  strings. The analysis should raise an error if you enter a non-existing value.
- Both inferences are set to run in `prior` and `posterior` modes
- You can enter arbitrary arguments to cmdstanpy's `CmdStanModel.sample` method
  in the `[sample_kwargs]` table.
- You can enter mode-specific overrides in `[mode_options.<MODE>]`. This can be
  handy if you want to run more or fewer iterations for a certain mode.

@@

```zsh
inferences
├── gpareto2006
│   ├── config.toml
│   └── idata
└── normal2006
    ├── config.toml
    └── idata
```

@@

Now that the inferences are ready it's time to check them out. Bibat provides a
Jupyter notebook at `notebooks/investigate.ipynb` for exactly this purpose. The
notebook's main job is to create plots and save them in the `plots` directory
when it is executed with the command `jupyter execute notebooks/investigate.ipynb`,
 which is the final step in the chain of commands that is triggered by `make
analysis`.

@@

The code from the example analysis's notebook for loading `InferenceData` was
reusable with a few tweaks to avoid missing file errors, so I kept it. On the
other hand, I wanted to make some different plots from the ones in the example
analysis, including some that required loading prepared data. To check out
everything I did, see
[here](https://github.com/teddygroves/bibat/blob/main/bibat/example_projects/baseball/notebooks/investigate.ipynb).

@@

To model the more recent data, all I had to do was create some new inference
folders with appropriate `prepared_data` fields in their `config.toml`
files. For example, here is the `config.toml` file for the `gparetobdb`
inference:

@@

After running `make analysis` one more time, I went back to the notebook
`notebooks/investigate.ipynb` and made plots of both models' posterior 1%-99%
success rate intervals for both datasets:

@@

The final step was to document my analysis. To do this I edited the file
`docs/report.qmd`, then ran `make docs`, which produced the
very HTML document that you are probably reading now! You can find the complete
`report.qmd` file
[here](https://github.com/teddygroves/bibat/blob/main/example_projects/baseball/docs/report.qmd).
